Resilience and Home-Space for WSTS \thanks{This work was partly done while the authors were supported by the Agence Nationale de la Recherche grant BraVAS (ANR-17-CE40-0028).}

### 2023-10-15 VMCAI-revised-version/final.tex
section

 A {\em well-quasi-ordering} (wqo) is any quasi-ordering $(X,\leq)$ such that, for any infinite sequence $x_0, x_1, x_2, ...$ in $X$, there exist indexes $i \leq j$ with $x_i \leq  x_j$. This property is equivalent to : every upward-closed set $\emptyset \neq U \subseteq X$ admits a finite basis $B \subseteq X$ such that $U=\uparrow B$.
%
%
%
Wqo admits many other equivalent formulations like : a qo $(X,\leq)$ is a wqo iff $(X,\leq)$ is well founded (i.e. there is no infinite strictly decreasing sequence
of elements of $X$) and $(X,\leq)$ contains no infinite antichains (an antichain is a subset of mutually incomparable elements of $X$). See other equivalences in \cite{schmitz:cel-00727025}. 
%
%
As an example, $(\N^d, \leq)$, the set of vectors of $d$ natural numbers (where $d$ is finite) with component-wise order is a wqo.
%
Quasi-orderings that have no 
infinite antichains
 enjoy a similar \emph{finite decomposition} property than wqo: every downward-closed subset $D \subseteq X$ can be decomposed into a \emph{finite} set of ideals $J_1, J_2, \ldots, J_n$ such that $D = \downarrow (J_1 \cup J_2 \cup \ldots \cup J_n)$.
%
%
In what follows, a downward-closed set $D$ is represented by its finite set of ideals (or by the minimal elements of its upward-closed complement), and an upward-closed set $U$ is represented by its finite set of minimal elements. \\

@@

\noindent
{\bf On effectivity.}
Let $\mathscr{S}=(S, \rightarrow, \leq)$ be a WSTS. We say that $\mathscr{S}$ is {\em effective} if there exists a pair of algorithms
($M_\rightarrow$, $M_\leq$) operating on $\N \times \N$ such that
$ M_\rightarrow$ computes the transition relation “$\rightarrow$” and 
$M_\leq$ the ordering relation “$\leq$”.
We say that $\mathscr{S}$ is {\em post-effective} if it is effective, and if there
exists an algorithm that computes $|\post(x)| \in \N \cup \{\infty\}$ 
on input $x \in X$. 
We say that $\mathscr{S}$ has {\em effective pred-basis} \cite{DBLP:journals/tcs/FinkelS01,DBLP:journals/iandc/AbdullaCJT00} if there exists an algorithm accepting
any state $s \in S$ and returning $pb(s)$, a finite basis of $\uparrow \pred(\uparrow s)$.
%
We say that $\mathscr{S}$
is {\em ideal-effective} \cite{BFM-ic17} if (1) the function mapping the encoding of a state $s$
to the encoding of the ideal $\downarrow s$ is computable; (2) inclusion of ideals is decidable; (3) the downward closure $\downarrow \post(I)$ expressed as a finite union of ideals is computable from the ideal $I$.
%

Now, we may recall a simple condition that insures that a finite basis of $\pred^*(\uparrow s )$ is computable for every $s \in S$.
We will use the following property: if $\mathscr{S}$ is a WSTS with strong compatibility and $U \subseteq S$ is upward-closed, then $\pred(U )$, $\pred^k(U )$ with $k\geq0$, and $\pred^*(U )$ are all upward-closed \cite{DBLP:journals/tcs/FinkelS01}.
For a WSTS $\mathscr{S}=(S, \rightarrow, \leq)$ and an upward-closed set $U  \subseteq S$, let us study the convergence of the sequence defined by $U_0=U$ and $U_k= U_{k-1} \cup \pred(U_{k-1})$ for $k \geq 1$. When $\mathscr{S}$ has strong compatibility, the sets $U_k$ are upward-closed and $U_k \subseteq U_{k+1}$ so we know that the sequence $(U_k)_k$ converges. Let us define the \emph{index} of convergence of the sequence $U_k$ as the smallest $k_0$ s.t. $U_k = U_{k_0}$ for all $k \geq k_0$. We may compute $k_0$ and we then have:  $\pred^*(U) = U_{k_0}$. When $\mathscr{S}$ have upward-compatibility but not strong compatibility, we can similarly compute $\pred^*(U)$ by studying the sequence $(\uparrow U_k)_k$ instead.
With the effective pred-basis hypothesis, we obtain:

@@

Let us recall that the \emph{completion}  \cite{BFM-ic17} of a WSTS $\mathscr{S}=(S,\rightarrow, \leq)$ is the associated ordered transition system $\hat{\mathscr{S}}=(Ideals(S),\rightarrow, \subseteq)$ where $Ideals(S)$ is the set of ideals of $S$ and $I \rightarrow J$ if $J$ belongs to the finite ideal decomposition of $\downarrow \post_{\mathscr{S}}(I)$. The completion is always finitely branching but it is not necessarly a WSTS since $\subseteq$ is not necessarly a wqo. $\hat{\mathscr{S}}$ is WSTS iff $\mathscr{S}=(S,\rightarrow, \leq)$ is $\omega^2$-WSTS. 
A WSTS $\mathscr{S}=(S,\rightarrow,\leq)$ is {\em completion-post-effective} if
it is post-effective, and there exists an algorithm $M_\downarrow$ that computes 
 $\downarrow s$, on input $s \in S$, and some algorithm $M_{\uparrow^C}$ that computes the ideal decomposition of $S \setminus \uparrow \{ s_1, s_2, \ldots, s_m\}$, on input
 $s_1, s_2, \ldots, s_m \in S$.
Coverability is shown decidable  [Theorem 44] in \cite{BFM-ic17} for completion-post-effective $\omega^2$-WSTS (we don't need the pred-basis hypothesis).\\

>>>>

 A {\em well-quasi-ordering} (wqo) is any quasi-ordering $(X,\leq)$ such that, for any infinite sequence $x_0, x_1, x_2, ...$ in $X$, there exist indexes $i \leq j$ with $x_i \leq  x_j$. This property is equivalent to : every upward-closed set $\emptyset \neq U \subseteq X$ admits a finite basis $B \subseteq X$ such that $U=\uparrow B$.
%
%
%
Wqo admits many other equivalent formulations like : a qo $(X,\leq)$ is a wqo iff $(X,\leq)$ is well founded (i.e. there is no infinite strictly decreasing sequence
of elements of $X$) and $(X,\leq)$ contains no infinite antichains (an antichain is a subset of mutually incomparable elements of $X$). See other equivalences in \cite{schmitz:cel-00727025}. 
%
%
As an example, $(\N^d, \leq)$, the set of vectors of $d$ natural numbers (where $d$ is finite) with component-wise order is a wqo.\\
Quasi-orderings that have no 
infinite antichains
 enjoy a similar \emph{finite decomposition} property than wqo: every downward-closed subset $D \subseteq X$ can be decomposed into a \emph{finite} set of ideals $J_1, J_2, \ldots, J_n$ such that $D = \downarrow (J_1 \cup J_2 \cup \ldots \cup J_n)$ See for example \cite{BFM-ic17}.
%
%
In what follows, a downward-closed set $D$ is represented by its finite set of ideals (or by the minimal elements of its upward-closed complement), and an upward-closed set $U$ is represented by its finite set of minimal elements. \\

@@

We introduce five types of effectivity that will be necessary to decide some resilence problems for differents models of WSTS.\\
\noindent
{\bf On effectivity.}
Let $\mathscr{S}=(S, \rightarrow, \leq)$ be a WSTS. We say that $\mathscr{S}$ is 
\begin{itemize}
\item {\em effective} if there exists a pair of algorithms
($M_\rightarrow$, $M_\leq$) operating on $\N \times \N$ such that
$ M_\rightarrow$ computes the transition relation “$\rightarrow$” and 
$M_\leq$ the ordering relation “$\leq$”.
\item {\em post-effective} if $\mathscr{S}$ is effective, and if there
exists an algorithm that computes $|\post(x)| \in \N \cup \{\infty\}$ 
on input $x \in X$. 
\item {\em effective pred-basis} \cite{DBLP:journals/tcs/FinkelS01,DBLP:journals/iandc/AbdullaCJT00} if there exists an algorithm accepting
any state $s \in S$ and returning $pb(s)$, a finite basis of $\uparrow \pred(\uparrow s)$.
%
\item {\em ideal-effective} \cite{BFM-ic17} if (1) the function mapping the encoding of a state $s$
to the encoding of the ideal $\downarrow s$ is computable; (2) inclusion of ideals is decidable; (3) the downward closure $\downarrow \post(I)$ expressed as a finite union of ideals is computable from the ideal $I$.
\item {\em completion-post-effective} if
$\mathscr{S}$ is post-effective, and there exists an algorithm $M_\downarrow$ that computes 
 $\downarrow s$, on input $s \in S$, and some algorithm $M_{\uparrow^C}$ that computes the ideal decomposition of $S \setminus \uparrow \{ s_1, s_2, \ldots, s_m\}$, on input
 $s_1, s_2, \ldots, s_m \in S$.
\end{itemize}
\alain{le rappel ci-dessous est-il entièrement utile ?}
Now, we may recall a simple condition that insures that a finite basis of $\pred^*(\uparrow s )$ is computable for every $s \in S$.
We will use the following property: if $\mathscr{S}$ is a WSTS with strong compatibility and $U \subseteq S$ is upward-closed, then $\pred(U )$, $\pred^k(U )$ with $k\geq0$, and $\pred^*(U )$ are all upward-closed \cite{DBLP:journals/tcs/FinkelS01}.
For a WSTS $\mathscr{S}=(S, \rightarrow, \leq)$ and an upward-closed set $U  \subseteq S$, let us study the convergence of the sequence defined by $U_0=U$ and $U_k= U_{k-1} \cup \pred(U_{k-1})$ for $k \geq 1$. When $\mathscr{S}$ has strong compatibility, the sets $U_k$ are upward-closed and $U_k \subseteq U_{k+1}$ so we know that the sequence $(U_k)_k$ converges. Let us define the \emph{index} of convergence of the sequence $U_k$ as the smallest $k_0$ s.t. $U_k = U_{k_0}$ for all $k \geq k_0$. We may compute $k_0$ and we then have:  $\pred^*(U) = U_{k_0}$. When $\mathscr{S}$ have upward-compatibility but not strong compatibility, we can similarly compute $\pred^*(U)$ by studying the sequence $(\uparrow U_k)_k$ instead.
With the effective pred-basis hypothesis, we obtain:

@@

Let us recall that the \emph{completion}  \cite{BFM-ic17} of a WSTS $\mathscr{S}=(S,\rightarrow, \leq)$ is the associated ordered transition system $\hat{\mathscr{S}}=(Ideals(S),\rightarrow, \subseteq)$ where $Ideals(S)$ is the set of ideals of $S$ and $I \rightarrow J$ if $J$ belongs to the finite ideal decomposition of $\downarrow \post_{\mathscr{S}}(I)$. The completion is always finitely branching but it is not necessarly a WSTS since $\subseteq$ is not necessarly a wqo. $\hat{\mathscr{S}}$ is WSTS iff $\mathscr{S}=(S,\rightarrow, \leq)$ is $\omega^2$-WSTS. 
%A WSTS $\mathscr{S}=(S,\rightarrow,\leq)$ is {\em completion-post-effective} if
%it is post-effective, and there exists an algorithm $M_\downarrow$ that computes 
% $\downarrow s$, on input $s \in S$, and some algorithm $M_{\uparrow^C}$ that computes the ideal decomposition of $S \setminus \uparrow \{ s_1, s_2, \ldots, s_m\}$, on input
% $s_1, s_2, \ldots, s_m \in S$.
Coverability is shown decidable  [Theorem 44] in \cite{BFM-ic17} for completion-post-effective $\omega^2$-WSTS (we don't need the pred-basis hypothesis).\\
